---
title: "Explanatory IRT for Dichotomous Data"
author: "Okan Bulut"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
description: >
  How does eirm can be used for estimating dichotomous and polytomous 
  explanatory IRT models in R?
vignette: >
  %\VignetteIndexEntry{Explanatory IRT for Dichotomous Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo = TRUE, eval = TRUE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)
options(width = 800)
library("eirm")
```

The following example shows how to use item-related and person-related explanatory variables to explain dichotomous responses in the verbal aggression data set. 

```{r, echo = TRUE, eval = TRUE}
data("VerbAgg")

mod2a <- eirm(formula = "r2 ~ -1 + situ + btype + mode + (1|id)", data = VerbAgg)

print(mod2a)
```

It is possible to visualize the parameters using an item-person map using `plot(mod2a)`, which returns the following plot:

```{r, echo = TRUE, eval = TRUE}
plot(mod2a)
```

In addition to the item-person map, we can also visualize the marginal effects in the model using the `marginalplot` function. This plot uses the [ggeffects](https://strengejacke.github.io/ggeffects/index.html) package to calculate the marginal effects and the [ggplot2](https://ggplot2.tidyverse.org/) package to create a plot. The following code will return a marginal effect plot with the three explanatory variables in `mod2a`.

```{r, echo = TRUE, eval = TRUE}
marginalplot(mod2a, predictors = c("situ", "btype", "mode"))
```

Lastly, we can compare nested explanatory models with each other. The following example shows the estimation of a more compact version of `mod2a` with one less variable and compares the two models (i.e., `mod2a` vs. `mod2b`).

```{r, echo = TRUE, eval = TRUE}
mod2b <- eirm(formula = "r2 ~ -1 + situ + btype + (1|id)", data = VerbAgg)

anova(mod2a$model, mod2b$model)
```
