---
title: "Simulating data for explanatory IRT models"
output: rmarkdown::html_vignette
description: >
  This vignette describes how to simulate response data with item-level and person-level covariates.
vignette: >
  %\VignetteIndexEntry{Simulating data for explanatory IRT models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, echo = TRUE, eval = TRUE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
```

### Linear Logistic Test Model (LLTM)

The Linear Logistic Test Model (LLTM) includes item-level covariates that can explain the difficulty level of items. Using the `simEIRM` function, we can create a categorical, item-level covariate with two levels. We will simulate response data with 20 items and 1000 persons. The first 10 items will be `itemtype = 0` and the second 10 items will be `itemtype = 1`. We will assume that the items with `itemtype = 1` are 0.5 logit more difficult on average. 

```{r, echo=FALSE}
library("eirm")
```

```{r}


data.lltm <- simEIRM(nitem = 20, nperson = 1000, itemtype.size = c(10, 10), 
                     itemtype.effect = c(0, 0.5), seed = 1234)

mod.lltm <- eirm("response ~ -1 + itemtype + (1 | person)", data = data.lltm)

print(mod.lltm)
```

